Bootstrapping Multi-atlas Hippocampal Segmentation with MAGeT

### 2013-02-13 paper.Rnw
Updated intro and methods

Automated segmentation techniques do not require human intervention but do
require {\it a priori} anatomical information to guide segmentations.  In this
paper we focus on methods that use manually segmented atlases as anatomical
priors, as these methods achieve some of the best automated hippocampal
segmentation accuracies to-date. This technique was pioneered by []\citep{} with
single-atlas (or model-based) segmentation.  Volumetric image registration is
used to estimate a fit between the neuroanatomy of an atlas and subject images.
Labelling of the subject image is achieved by applying the resulting
transformation to the atlas labels to bring them into the subject image space
(\{em label propagation}). This method is limited in accuracy by the
introduction of {methodological} errors in registration and partial volume
effects in label resampling, and {conceptual} errors introduced when the anatomy
of the atlas is not representative of the subject.

Multi-atlas segmentation techniques address these limitations by combining
segmentation information from a series of expertly segmentated atlases
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2012,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a subject image, and label propagation is
performed to produce several labellings of the subject image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used merge
these labels into a definitive segmentation for the subject.

{\em Atlas selection} techniques are often used to filter out atlases dissimilar
to a subject image in order to reduce {conceptual} errors \citep{Aljabar2009}.

@@

In \citep{Lotjonen2012}, the authors use a proprietary non-linear registration
method based on intensity differences, and post-processing step taking into
account these intensity differences to 
- TODO: explain "tidy up"
tidy up
fused segmentations. Using images from the ADNI1 baseline dataset, they achieved
a mean DSC of 0.885 with 30 atlases.

The LEAP algorithm, Wolz et al.  is an elegant modification to the basic
multi-atlas strategy: the atlas library is grown, beginning with a set of
manually labelled atlases and successively incorporating unlabelled subject
images after being labelling using multi-atlas techniques\citep{Wolz2010}.  The
sequence in which subject images are labelled is chosen so that the similarity
between the atlas images and the target images is minimised at each step,
effectively allowing for deformations between very dissimilar images to be
broken up into sequences of smaller deformations.  With an atlas library of 30
MR images, LEAP was used to segment the ADNI1 baseline dataset, achieving a mean
Dice score of 0.85 with manual segmentations.

- TODO: "probablistic methods?"

@@

Aside from the algorithmic choices used in multi-atlas segmentation, it is
natural to ask about how the features of the atlases themselves impact the
resulting segmentations. As noted, by choosing atlases ranked most similar to a
subject image by voxel intensity profile segmentation accuracy is improved,
suggesting that neuroanatomical similarity plays a role\citep{Aljabar2009}.
Carmichael et al. explored this directly and found that when using only one
atlas the important factors leading to improved accuracy are that the atlas have
neuroanatomical features that match the subject, and that the atlas segmentation
use the same protocol as the gold-standard \citep{Carmichael2005}. Nestor et al.
found that hippocampal segmentation protocols that include more dorsal
white-matter and posterior anatomy tended to produce higher overlap and better
accuracy at distinguishing disease classes in the ADNI1 1 year dataset
\citep{Nestor2012}. These results suggest both atlas library neuroanatomy and
delineation protocol play a significant role in the resulting segmentation.

@@

Our algorith, called MAGeT brain ({\em M}ultiple {\em A}utomatically {\em
Ge}nerated {\em T}emplates), is an extension to the basic multi-atlas-based
segmentation schema\citep{Chakravarty2012}. Principally, we explore the
possibility of using a small atlas library to bootstrap a much larger
\emph{template library} composed of images taken from the target population.
The template library is then used to segment the subjects in a similar fashion
to basic multi-atlas segmentation: by label propagation and label fusion. The
intuition driving this approach is that by generating a template library we
leverage the unique neuroanatomy of subject population on hand to initialize the
segmentation process and improve accuracy over direct propagation from the atlas
library to unlabelled subjects while also using fewer manually segmented
atlases.

The insight of generating a template library is not new.  Heckemann et al.
compared ``indirect'' segmentation -- taking a single atlas and propagating the
labels to intermediate subjects before fusing them in a target image space -- to
multi-atlas segmentation and found that the indirect approach performed worse
\citep{Heckemann2006}. In this paper we continue the same line of investigation
but explore the performance when using multiple atlases as well as the effect of
different registration and fusion methods.

@@

In previous work \citep{Chakravarty2012}, we applied MAGeT brain to segmentation
of the human striatum, globus pallidus, and thalamus using a single
histologically-derived atlas. The contribution of the present work is to extend
our approach to the human hippocampus and perform a rigorous validation. We
first conduct an intensive cross-validation of MAGeT on a pool of 69 MRIs from
the ADNI database, comparing the performance of MAGeT against basic multi-atlas
segmentation using a range of atlas and template library sizes, registration and
fusion methods.  We then compare MAGeT segmentations of the entire ADNI
screening dataset with those of established automated and manual segmentation
techniques.  Finally, we compare MAGeT segmentations to those of manual
segmentations from a dataset of first episode schizophrenic patient MRIs.

@@

\subsection{MAGeT Brain Algorithm}
We will use the term {\it atlas} to refer to two image volumes: an intensity
image ({\it atlas image}) and a corresponding manual segmentation image ({\it
atlas labels}). A {\it subject} image is an intensity image for which we would
like an segmentation.  The term {\it template} will be used to refer more
generically to any image and corresponding labelling, manual or computed, when
it is {\it used as} as a model in the segmentation of another image. In other
publications, the terms atlas and template are used interchangeably, but in this
paper we force the distinction to make it clear when an image has a manual
segmentation.  The terms {\it atlas library} and {\it template library} mean a
set of such images.

The simplest form of multi-atlas segmentation combines labellings derived from
several atlases by way of label fusion \citep{Heckemann2006,Heckemann2011}. We
will refer to this as {\em basic multi-atlas segmentation}, and specific the
basic schema as:

    1. an atlas library and set of subject images are given as
       input, and the atlas library is used as a template library in the
       following steps; 
    2. each atlas intensity image is nonlinearly registered to each subject
       intensity image; 
    3. label images from each atlas are propagated via the resulting
       transformations to the subject image space; and 
    4. the resulting labels are fused to produce a single, definitive
       segmentation.

The particular registration and voting method used are left unspecified.

MAGeT brain is best understood as an extension of
the basic multi-atlas segmentation schema. An additional stage is included in
which the template library is constructed by labelling a subset of subject
images, rather than by using the atlas library. The choice of subjects used in
the template library can be made to reflect the neuroanatomy or demographics of
the subject set as a whole (for instance, by sampling equally from cases and
controls).

Labels from each atlas image are propagated to
each template library image via the transformation resulting from a non-linear
registration between pair of images. As a result, each template library image
has a label from each atlas. Basic multi-atlas segmentation is then used to
produce segmentations for the entire set of unlabelled images (including those
images used in the template library). Figure \ref{alg:MAGeT} describes the MAGeT
brain algorithm in pseudocode.  Source code can be found at
\url{http://github.com/pipitone/MAGeTbrain}.

\begin{algorithm}
  \scriptsize
  \caption{Pseudocode for the MAGeT Brain algorithm}
  \label{alg:MAGeT}
  \begin{algorithmic}
    \Function{BasicMultiAtlasSegmentation}{Templates, Subjects}
      \ForAll{$subject$}
        \ForAll{$template$}
          \State propagate all labels for template to subject space
          \State store subject labels
        \EndFor
        \State fuse subject labels
      \EndFor
    \EndFunction
    \\
    \Function {MAGeTBrain}{Subjects, Atlases, n}
      \For{$i = 1 \to n$}
        \State choose a subject to be used as a template
        \State propagate labels from each atlas to template space
        \State store the template with all of its labels
      \EndFor
      \State MultiAtlas(Templates, Subjects)
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\begin{figure}[h]
  \centering
  \includegraphics[width=0.5\textwidth]{figure/MAGeT-figure.png}
  \caption{Diagram of the MAGeT Brain algorithm \label{fig:MAGeT-diagram}}
\end{figure}

@@

%%%%%                       Image pre-processing                           %%%%%

\subsection{Image pre-processing}

Before images were registered, the N3 algorithm \citep{Sled1998} is first used
to minimize the intensity nonuniformity in each of the atlases and unlabeled
subject images.

%%%%%                       Registration Methods                           %%%%%

\subsection{Registration}

@@

Label fusion is a term given to the process of combining the information from
several candidate labellings for an MR image into a single labelling.  In this
paper we explore the benefits of three different fusion methods.

@@

Labels are propagated from all template library images to a subject.  Each
output voxel is given the most frequent label at that voxel location amongst
all candidate labellings.  Ties are broken arbitrarily.

@@

An optimal combination of subjects from the template library has previously been
shown to improve segmentation accuracy \citep{Aljabar2009,Collins2010}.  In this
method, each template library image is ranked in similarity to each unlabelled 
image by the normalized cross-correlation (CC) of image intensities after linear
registration, over a region of interest (ROI) generously encompassing the 
hippocampus.  Only the top ranked template library image labels are used in a
voxel-wise majority vote. The ROI is heuristically defined as the extent of all
atlas labels after linear registration to the template, dilated by three voxels
\citep{Chakravarty2012}.  The number of top ranked template library image labels
is a configurable parameter.

@@

The agreement between two segmentations can assessed using the Dice Similarity
Coefficient (DSC):

\begin{equation*} 
DSC=\frac{2|A \cap B|}{|A| + |B|}
\end{equation*}

where $A$ and $B$ are the regions being compared, and the cardinality is the
volume measured in voxels. Using this measure we can compare an automatically
generated segmentation to a gold-standard segmentation.

@@

%%%%%
%%%%%               Experiments: ADNI-1 Screening Validation               %%%%%
%%%%%
\subsubsection{ADNI-1 Screening Validation}
\todo{comparison across field strengths, a la Heckemann2011?}
To test the accuracy of MAGeT brain on a real-world task we segment the entire
ADNI-1 dataset using an atlas set that is not representative of the subject set.

@@

\todo{Can we statistically capture what "peak" performance is?  Something like,
the point at which gains become statistically insignificant?}

>>>>

Automated segmentation techniques do not require human intervention but do
require {\it a priori} anatomical information to guide segmentations.  In this
paper we focus on methods that use manually segmented MRI atlases as anatomical
priors, as these methods achieve some of the best automated hippocampal
segmentation accuracies to-date. This technique was first developed using a
single atlas prior (known as single-atlas, or model-based,
segmentation)\citep{Haller1997,Csernansky1998}.  Volumetric image registration
is used to estimate a fit between the neuroanatomy of an atlas and target
images.  Labelling of the target image is achieved by applying the resulting
transformation to the atlas labels to bring them into the target image space
(\{em label propagation}). This method is limited in accuracy by the
introduction of estimation errors in registration and partial volume effects in
label resampling, and errors introduced when the anatomy of the atlas is
unrepresentative of the target anatomy.

Multi-atlas segmentation techniques address these limitations by combining
segmentation information from a series of expertly segmented atlases
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2012,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used merge
these labels into a definitive segmentation for the target.

@@

In \citep{Lotjonen2012}, the authors use a proprietary non-linear registration
method based on intensity differences, and post-processing step using a graph
cuts or EM approach with intensity spatial prior models to optimise the fused
segmentations.  Using images from the ADNI1 baseline dataset and 30 atlases.

{\em Atlas selection} techniques are often used to filter out atlases dissimilar
to a target image in order to reduce {conceptual} errors \citep{Aljabar2009}.

The LEAP algorithm, Wolz et al.  is an elegant modification to the basic
multi-atlas strategy: the atlas library is grown, beginning with a set of
manually labelled atlases and successively incorporating unlabelled target
images after being labelling using multi-atlas techniques\citep{Wolz2010}.  The
sequence in which target images are labelled is chosen so that the similarity
between the atlas images and the target images is minimised at each step,
effectively allowing for deformations between very dissimilar images to be
broken up into sequences of smaller deformations.  With an atlas library of 30
MR images, LEAP was used to segment the ADNI1 baseline dataset, achieving a mean
Dice score of 0.85 with manual segmentations.

%- TODO: "probablistic methods?"

@@

Aside from the algorithmic choices used in multi-atlas segmentation, it is
natural to ask about how the features of the atlases themselves impact the
resulting segmentations. As noted, by choosing atlases ranked most similar to a
target image by voxel intensity profile segmentation accuracy is improved,
suggesting that neuroanatomical similarity plays a role\citep{Aljabar2009}.
Carmichael et al. explored this directly and found that when using only one
atlas the important factors leading to improved accuracy are that the atlas have
neuroanatomical features that match the target, and that the atlas segmentation
use the same protocol as the gold-standard \citep{Carmichael2005}. Nestor et al.
found that hippocampal segmentation protocols that include more dorsal
white-matter and posterior anatomy tended to produce higher overlap and better
accuracy at distinguishing disease classes in the ADNI1 1 year dataset
\citep{Nestor2012}. These results suggest both atlas library neuroanatomy and
delineation protocol play a significant role in the resulting segmentation.

@@

Our algorith, called MAGeT brain ({\em M}ultiple {\em A}utomatically {\em
Ge}nerated {\em T}emplates), is an extension to the basic multi-atlas-based
segmentation schema\citep{Chakravarty2012}. Principally, we explore the
possibility of using a small atlas library to bootstrap a much larger
\emph{template library} composed of images taken from the target population.
The template library is then used to segment the targets in a similar fashion
to basic multi-atlas segmentation: by label propagation and label fusion. The
intuition driving this approach is that by generating a template library we
leverage the unique neuroanatomy of target population on hand to initialize the
segmentation process and improve accuracy over direct propagation from the atlas
library to unlabelled targets while also using fewer manually segmented
atlases.

The insight of generating a template library is not new.  Heckemann et al.
compared ``indirect'' segmentation -- taking a single atlas and propagating the
labels to intermediate targets before fusing them in a target image space -- to
multi-atlas segmentation and found that the indirect approach performed worse
\citep{Heckemann2006}. In this paper we continue the same line of investigation
but explore the performance when using multiple atlases as well as the effect of
different registration and fusion methods.

@@

In previous work \citep{Chakravarty2012}, we applied MAGeT brain to segmentation
of the human striatum, globus pallidus, and thalamus using a single
histologically-derived atlas. The contribution of the present work is to extend
our approach to the human hippocampus and perform a series of experiments to
rigorous validate the method. First, we conduct an extensive cross-validation of
MAGeT and basic multi-atlas segmentation on a subset of the ADNI1 dataset to
assess the accuracy of MAGeT brain under various parameter settings (number of
atlases and templates, registration and fusion methods). With the best
performing parameter configuration discovered above, we estimate MAGeT
intra-rater reliability by segmenting separately acquired T1 images of the
atlas subjects. For this experiment, we use the Winterburn atlases: digital
hippocampal subfield segmentations of five in-vivo high-resolution (300u
isotropic) T1-weighted MR scans\citep{Winterburn2013}. To validate MAGeT in a
real world situation, we segment the entire ADNI1 screening dataset and compare
our segmentations to established automated and manual segmentations.
Additionally, to ensure MAGeT brain accuracy across disease categories, we also
compare MAGeT segmentations to manual segmentations of 139 first episode
schizophrenic patients.

@@

\subsection{MAGeT Brain Algorithm}
In this paper we will be making a distinction between an atlas and a template --
typically these terms are used roughly interchangeably. The term {\it atlas} is
taken to refer to two image volumes: an intensity image ({\it atlas image}) and
a corresponding manual segmentation image ({\it atlas labels}). {\it Template}
refers more generically to any image and corresponding labelling, manual or
computed, when it is {\it used as} as a model in the segmentation of another
image.  The terms {\it atlas library} and {\it template library} mean a set of
such images. Additionally, we will use the terms {\it target} to refer to an
intensity image for which we would like an segmentation.

The simplest form of multi-atlas segmentation combines labellings derived from
several atlases by way of label fusion \citep{Heckemann2006,Heckemann2011}. We
will refer to this as {\em basic multi-atlas segmentation}. The basic schema as
for this method is as follows: 
\begin{itemize}
\item   An atlas library and set of target images are given as input. The atlas
        library is used as a template library in the following steps; 
\item   Each atlas intensity image is nonlinearly registered to each target 
        intensity image; 
\item   label images from each atlas are propagated via the resulting
        transformations to the target image space; and 
\item   the resulting labels are fused to produce a single, definitive
        segmentation. 
\end{itemize}       
The particular registration and voting method used are left unspecified.

MAGeT brain is best understood as an extension of the basic multi-atlas
segmentation schema. Additional stages construct the template library by
labelling a subset of target images, rather than by using the atlas library.
The choice of targets used in the template library can be made to reflect the
neuroanatomy or demographics of the target set as a whole (for instance, by
sampling equally from cases and controls). A truncated version of basic
multi-atlas segmentation is used to label the template library images; label
fusion is not performed. Instead, each template image gets a labelling from each
atlas image. A second round of basic multi-atlas segmentation uses the template
library to segment the entire set of target images (including those images used
in the template library). Label fusion in this final step fuses all labels from
all templates.  Figure \ref{alg:MAGeT} describes the MAGeT brain algorithm in
pseudocode.

Source code for MAGeT brain can be found at
\url{http://github.com/pipitone/MAGeTbrain}.

\begin{algorithm}
  \scriptsize
  \caption{Pseudocode for the MAGeT Brain algorithm}
  \label{alg:MAGeT}
  \begin{algorithmic}
    \Function{BasicMultiAtlasSegmentation}{Templates, Subjects}
      \ForAll{$target$}
        \ForAll{$template$}
          \State propagate all labels for template to target space
          \State store target labels
        \EndFor
        \State fuse target labels
      \EndFor
    \EndFunction
    \\
    \Function {MAGeTBrain}{Subjects, Atlases, n}
      \For{$i = 1 \to n$}
        \State choose a target to be used as a template
        \State propagate labels from each atlas to template space
        \State store the template with all of its labels
      \EndFor
      \State MultiAtlas(Templates, Subjects)
    \EndFunction
  \end{algorithmic}
\end{algorithm}

%\begin{figure}[h]
%  \centering
%  \includegraphics[width=0.5\textwidth]{figure/MAGeT-figure.png}
%  \caption{Diagram of the MAGeT Brain algorithm \label{fig:MAGeT-diagram}}
%\end{figure}

@@

c
%%%%%                       Registration Methods                           %%%%%

\subsection{Registration Methods}

Before registration, all images undergo preprocessing with the N3 algorithm
\citep{Sled1998} to minimize intensity nonuniformity. In our experiments we use
one of two non-linear image registration methods.

@@

Label fusion is a term given to the process of combining the information from
several candidate labellings for an intensity image into a single labelling.  In
this paper we explore three different fusion methods.

@@

Labels are propagated from all template library images to a target.  Each
output voxel is given the most frequent label at that voxel location amongst
all candidate labellings.  Ties are broken arbitrarily.

@@

An optimal combination of targets from the template library has previously been
shown to improve segmentation accuracy \citep{Aljabar2009,Collins2010}.  In this
method, each template library image is ranked in similarity to each unlabelled 
image by the normalized cross-correlation (CC) of image intensities after linear
registration, over a region of interest (ROI) generously encompassing the 
hippocampus.  Only the top ranked template library image labels are used in a
voxel-wise majority vote. The ROI is heuristically defined as the extent of all
atlas labels after linear registration to the template, dilated by three voxels
\citep{Chakravarty2012}.  The number of top ranked template library image labels
is a configurable parameter.

@@

The Dice similarity index (SI) assesses the agreement between two segmentations.
It is one of the most widely used measures of segmentation performance, and we
use it as the basis of comparison throughout this paper.  Additionally, we
report other commonly used measures:

\begin{itemize}
 \item \begin{equation*}Similarity index (SI) =\frac{2|A \cap B|}{|A| + |B|}\end{equation*}
 \item \begin{equation*}Jaccard (J) =\frac{2|A \cap B|}{|A| + |B|}\end{equation*}

where $A$ and $B$ are the regions being compared, and the cardinality is the
volume measured in voxels. .

@@

%%%%%
%%%%%               Experiments: ADNI-1 Screening Validation               %%%%%
%%%%%
\subsubsection{ADNI-1 Screening Validation}
\todo{comparison across field strengths, a la Heckemann2011?}
To test the accuracy of MAGeT brain on a real-world task we segment the entire
ADNI-1 dataset using an atlas set that is not representative of the target set.

@@

- True for all parameter settings! (so works generally..)
\todo{Can we statistically capture what "peak" performance is?  Something like,
the point at which gains become statistically insignificant?}
